Assessing landscape N removal in coastal New England catchments using the

### 2023-06-08 manuscript/manuscript.Rmd
Updates based on EPA reviews done.

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with a new R package, nsink"
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features within watersheds can act as landscape
  sinks for N, such as wetlands, streams and ponds that transform dissolved N into
  gaseous N, effectively removing it from the aquatic system. Identifying and evaluating
  these landscape sinks and their spatial relationship to N sources can help managers
  understand the effects of alternative decisions on downstream resources. N-Sink
  is an approach that uses widely available GIS data to identify landscape sinks within
  HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
  of those sinks on N movement through the catchment. A series of static maps are
  produced to visualize N removal efficiency, transport and delivery, the latter in
  the form of an index. To facilitate data acquisition, processing and visualization,
  an R package `nsink` was developed and applied to produce static maps for all HUC-12
  catchments in southern New England. The R package can be used to develop N delivery
  maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
  website to explore the already mapped areas. Users can also investigate specific
  flowpaths interactively by clicking on any location within the catchment. A flowpath
  is generated, along with a table describing N removal along each segment."
---

@@

Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal]. Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland]. Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations. N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N. This watershed flow path approach provides a means to easily visualize the relative sensitivity of an area within a watershed that may require more aggressive N management at the source, protection of existing N retentive areas, or restoration and construction of downgradient N retentive areas that are needed to reduce N delivery to estuaries.

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time within landscape features that support N removal is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] . Datasets used by `nsink` include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web]. The static maps illustrate 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed. This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and c) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream. N-Sink and `nsink` integrate these data to make the application of complex hydrologic and biogeochemical processes accessible to end users and stakeholders in a visually clear and useful format. Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.

@@

## Using the `nsink` package
The `nsink` package provides several functions (Table 1) to set up and run an N-Sink analysis for a specified 12-digit HUC. By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.  Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC. The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.

@@

## Download the data
The first step for an N-Sink analysis with the `nsink` package is to download the required datasets.  The only required information to download the data is a HUC identifier.  The `nsink` package was developed using 12-digit HUC IDs, but larger HUCs (e.g. 8-digit) may also be used.  There are two functions provided for downloading the data.

`nsink_get_huc_id()` - search 12-digit HUC IDs using a known location name.

`nsink_get_data()` - download and save, in a specified folder, hydrography, soils, and land cover data for the specified 12-digit HUC. These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC. The `nsink` package utilizes widely available data from several U.S. Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.  The HUC ID is required and users may specify a path for storing the data as well as indicate whether or not to download the data again if they already exist in the data directory.

@@

- raster_method: Contains a raster-based approach to calculating removal.  Used for the static maps showing potential N removal within landscape sinks.
- land_removal: Represents land-based nitrogen removal within vegetated hydric soils (i.e., with impervious surface/developed areas removed).
- network_removal: Contains removal along the NHD Plus flow network. Removal is calculated separately for streams and waterbodies (i.e., lakes and reservoirs).

@@

SHOULD WE ADD SOMETHING ABOUT AREAS WHERE GROUNDWATER FLOW DOMINATES (E.G. CAPE COD)?

@@

The maps produced by N-Sink can be helpful in prioritizing source controls, sink protection and/or restoration. For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions. By better understanding the important role that N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters. Variances granted on a case-by-case basis can quickly undermine local water quality goals. Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize land acquisition or wetland restoration projects.

>>>>

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with a new R package, nsink"
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features within watersheds can act as landscape
  sinks for N, such as wetlands, streams and ponds that transform dissolved N into
  gaseous N, effectively removing it from the aquatic system. Identifying and evaluating
  these landscape sinks and their spatial relationship to N sources can help managers
  understand the effects of alternative decisions on downstream resources. N-Sink
  is an approach that uses widely available GIS data to identify landscape sinks within
  HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
  of those sinks on N movement through the catchment. A series of static maps are
  produced to visualize N removal efficiency, transport and delivery, the latter in
  the form of an index. To facilitate data acquisition, processing and visualization,
  an R package `nsink` was developed and applied to produce static maps for all HUC-12
  catchments in southern New England. The R package can be used to develop N delivery
  maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
  website to explore previously mapped areas. Users can also investigate specific
  flowpaths interactively by clicking on any location within the catchment. A flowpath
  is generated, along with a table describing N removal along each segment."
---

@@

Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal]. Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland]. Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations. N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N [@hollister2022nsink; @R]. This watershed flow path approach provides a means to easily visualize the relative sensitivity of an area within a watershed that may require more aggressive N management at the source, protection of existing N retentive areas, or restoration and construction of downgradient N retentive areas that are needed to reduce N delivery to estuaries.

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time within landscape features that support N removal through denitrification is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] . Datasets used by `nsink` include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web]. The static maps illustrate 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed. This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and 3) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, after accounting for the potential N removal as water moves downstream. N-Sink and `nsink` integrate these data to make the application of complex hydrologic and biogeochemical processes accessible to end users and stakeholders in a visually clear and useful format. Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.

<!-- EPA Review Comment: If the paper is about N-Sink integrating nsink it feels like the introduction could benefit from some more mention of N-Sink’s value-add at the end of the intro. Even just a sentence about a pre-analyzed web map being accessible to people who aren’t familiar with R.-->

@@

## Using the `nsink` package
The `nsink` package provides several functions (Table 1) to set up and run an N-Sink analysis for a specified 12-digit HUC. By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.  Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC. The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Index, and N Delivery Index.

@@

## Download the data
The first step for an N-Sink analysis with the `nsink` package is to download the required datasets.  The only required information to download the data is a HUC identifier.  The `nsink` package was developed using 12-digit HUC IDs from NHDPlus, but larger HUCs (e.g. 8-digit) may also be used.  There are two functions provided for downloading the data.

- `nsink_get_huc_id()` - search 12-digit HUC IDs using a known location name.

- `nsink_get_data()` - download and save, in a specified folder, hydrography, soils, and land cover data for the specified 12-digit HUC. These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC. The `nsink` package utilizes widely available data from several U.S. Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.  The HUC ID is required and users may specify a path for storing the data as well as indicate whether or not to download the data again if they already exist in the data directory.

@@

- raster_method: Contains a raster-based approach to calculating removal.  Used for the static maps showing potential N removal within landscape sinks.
- land_removal: Represents land-based nitrogen removal within vegetated hydric soils with impervious surface removed).
- network_removal: Contains removal along the NHD Plus flow network. Removal is calculated separately for streams and waterbodies (i.e., lakes and reservoirs).

@@

<!-- SHOULD WE ADD SOMETHING ABOUT AREAS WHERE GROUNDWATER FLOW DOMINATES (E.G. CAPE COD)? -->

@@

The maps produced by N-Sink can helpl prioritize source controls, sink protection and/or restoration. For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions. By better understanding the important role that N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters. Variances granted on a case-by-case basis can quickly undermine local water quality goals. Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize land acquisition or wetland restoration projects.
